---
title: "gene_phen_analyses"
output: html_document
date: '2022-07-20'
---

#Loading Libraries that are used throughout code
```{r, load libraries}
install.packages("oddsratio")

library(oddsratio)
library(stringr)
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("biomaRt")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("clusterProfiler")

library(BiocManager)
library(clusterProfiler)
library(org.Hs.eg.db)

```
# Creating Vector Objects to Store ID,Name, & is_a
```{r, create vector objects}
current_hpo_node <- list(id = character(), name = character(), is_a = character())
hpo_list <- list()
```
# Processing obo file into R and creating hpo_list
```{r, process obo file}
con <- file("hp.obo.txt", "r")
while ( TRUE ) {
  line <- readLines(con, n = 1)
  if ( length(line) == 0 ) {
    hpo_list[[length(hpo_list)+1]] <- current_hpo_node
    break
  }
  
  if (grepl("^id: HP:",line)){
   
    if (length(current_hpo_node$id) != 0){
      hpo_list[[length(hpo_list)+1]] <- current_hpo_node
      current_hpo_node <- list(id = character(), name = character(), is_a = character())
    }
    
    current_hpo_node$id <- str_replace(line, "id: HP:", "")
  }
  if (grepl("^name",line)){
    current_hpo_node$name <- str_replace(line, "name:", "")
    
  }

  if (grepl("^is_a",line)){
    temp <-str_replace(line, "is_a: HP:", "")
    temp <- str_extract(temp,"[0-9]{7}")
    current_hpo_node$is_a <- c(current_hpo_node$is_a,
                               temp)
  }
}

close(con)

node_list = hpo_list 
```
#Defining Tree functions that will be used to build the tree and add a node to the tree
```{r, define tree funcs}
add_node_to_tree <- function(tree, node){
  if (node$id < tree$id){
    if (all(is.na(tree$left))){
      tree$left <- node
      
    }else {
      tree$left <- add_node_to_tree(tree$left, node)
      return(tree)
    }
  }
  if (node$id > tree$id){
    if (all(is.na(tree$right))){
      tree$right <- node
    }else {
      tree$right <- add_node_to_tree(tree$right, node)
      return(tree)
    }
  
  }
  if (node$id == tree$id){
    print('Warning: Current Node is equal to Tree Node')
  }
  return(tree)
}

build_tree <- function(node_list){
  which_node <- sample(1:length(node_list),1)
  node <- node_list[[which_node]]
  node_list[[which_node]] <- NULL
  node$left <- NA
  node$right <- NA
  tree <- node
  
  while(length(node_list)>0){
    which_node <- sample(1:length(node_list),1)
    node <- node_list[[which_node]]
    node$left <- NA
    node$right <- NA
    node_list[[which_node]] <- NULL
    tree <- add_node_to_tree(tree, node)
  }
  return(tree)
} 

```
#Building the actualy binary search tree
```{r, building binary search tree}
hpo_bin_search_tree <- build_tree(hpo_list)
```
#Set of functions that are able to pull a node from the tree, get a nodes ancestors, and add genes to a node
```{r, searching tree functions}
get_depth_of_node_in_tree = function(id, tree, depth=0){
  if (id < tree$id){
    if (all(is.na(tree$left))){
      return(-1)
    }
    return(get_depth_of_node_in_tree(id, tree$right, depth = depth+1))
  }
  if (id > tree$id){
    if (all(is.na(tree$right))){
      return(-1)
    }
    return(get_depth_of_node_in_tree(id, tree$right, depth = depth+1))
  }
  if (id == tree$id){
    return(depth)
  }
  else 
    return(get_depth_of_node_in_tree(id, tree$right, depth = depth+1))
}

get_node_from_tree <- function(id, tree){
  if (id < tree$id){
    if (all(is.na(tree$left))){
      return(-1)
    }
    return(get_node_from_tree(id, tree$left))
  }
  if (id > tree$id){
      if (all(is.na(tree$right))){
        return(-1)
      }
    return(get_node_from_tree(id, tree$right))
  }
  if (id == tree$id){
    tree$left = NULL
    tree$right = NULL
    node <- tree
    return(node)
  }
}

get_ancestors <- function(id,tree,ancestors = character(0)){
  node <- get_node_from_tree(id,tree)
  ancestors <- c(ancestors, node$id)
  if (length(node$is_a)== 0){
    return(ancestors)
  }
  for (is_a in node$is_a){
      ancestors <- unique(c(ancestors,get_ancestors(is_a, tree, ancestors)))
  }
  return(ancestors)
}

add_genes_to_node <- function(id, tree, gene_ids){
  if(id < tree$id){
    if(all(is.na(tree$left))){
      return(tree)
    }
    tree$left <- add_genes_to_node(id, tree$left, gene_ids)
    return(tree)
  }
  if(id > tree$id){
    if(all(is.na(tree$right))){
      return(tree)
    }
    tree$right <- add_genes_to_node(id, tree$right, gene_ids)
    return(tree)
  }
  if(id == tree$id){
    tree$genes_assorted <- unique(c(tree$genes_assorted, gene_ids))
    return(tree)
  }
}

```
#Reading in the file containing the genes linked to the HPO Id's
```{r,reading phen_to_gene file}
phen_to_gene <- read.csv("phenotype_to_genes.txt", sep = "\t")
phen_to_gene$HPO.id <- str_replace(phen_to_gene$HPO.id, "HP:", "")
unique_hpo <- unique(phen_to_gene$HPO.id)[1:length(unique(phen_to_gene$HPO.id))]
```
#Adding the genes to the search tree based on the HPO ID and Ancestor Id's
```{r,adding genes to the hpo_bin_search_tree}
for (i in 1:length(unique(phen_to_gene$HPO.id))){
  current_hpo_id <- unique(phen_to_gene$HPO.id)[i]
  current_hpo_id <- str_replace(current_hpo_id, "HP:", "")
  which <- which(phen_to_gene$HPO.id == current_hpo_id)
  temp_genes <- phen_to_gene[which,]$entrez.gene.symbol
  current_hpo_id_w_ancestors <- get_ancestors(current_hpo_id, hpo_bin_search_tree)
  for(j in 1:length(current_hpo_id_w_ancestors)){
    hpo_bin_search_tree <- add_genes_to_node(current_hpo_id_w_ancestors[j], hpo_bin_search_tree, temp_genes)
  }
  
}
```

```{r, inserting gene_ensemble}
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl", GRCh=37)
gene_ensemble <- getBM(attributes=c('chromosome_name','hgnc_symbol', 'ensembl_gene_id','gene_biotype'),
               filters = list('biotype'='protein_coding'),
               mart = ensembl, useCache = F)

```
```{r, GO enrichment}
node <- get_node_from_tree("0005948", hpo_bin_search_tree)

  temp_genes <- node$genes_assorted
  gene_subset <- gene_ensemble$ensembl_gene_id[which(is.element(gene_ensemble$hgnc_symbol, temp_genes))]
  ego <- enrichGO(gene = gene_subset,
                  universe = gene_ensemble$ensembl_gene_id,
                  keyType = "ENSEMBL",
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  pAdjustMethod = "BH",
                  qvalueCutoff = 0.05,
                  readable = TRUE,
                  pool=TRUE)


head(ego)
```
```{r,inseting chromatin modifiers }
chrom <- read.csv("chromatin_modifiers.csv")
chromatin_modifiers <- character()
chromatin_modifiers <- unique(chrom$hgnc_symbol) 
```
```{r, adding chromatin modifiers to search tree}
add_chrom_modifier_to_node <- function(tree, chrom_m){
    if(!all(is.na(tree$left))){tree$left <- add_chrom_modifier_to_node(tree$left, chrom_m)
  }
    if(!all(is.na(tree$right))){tree$right <- add_chrom_modifier_to_node(tree$right, chrom_m)
  }
  tree$chrom_assorted <- tree$genes_assorted[which(is.element(tree$genes_assorted, chromatin_modifiers))]
 return(tree)
}

hpo_bin_search_tree <- add_chrom_modifier_to_node(hpo_bin_search_tree, chromatin_modifiers)
```
```{r,adding fisher test to search tree }
add_fisher_to_node <- function(tree){
  if(!all(is.na(tree$left))){tree$left <- add_fisher_to_node(tree$left)
  }
  if(!all(is.na(tree$right))){tree$right <- add_fisher_to_node(tree$right)
  }
  if (length(tree$chrom_assorted) != 0){
    tree$fisher_test <- fisher.test(t(matrix(c(length(tree$chrom_assorted)
                                             ,length(chromatin_modifiers) - length(tree$chrom_assorted) 
                                             ,length(tree$genes_assorted) - length(tree$chrom_assorted)
                                             ,length(unique(gene_ensemble$ensembl_gene_id)) - length(tree$genes_assorted)),
                                           nrow=2)))
  }
  return(tree)
}

hpo_bin_search_tree <- add_fisher_to_node(hpo_bin_search_tree)
```
```{r, creating dataframe from search tree}
tree_to_dataframe <- function(tree, current_dataframe=NULL){

  dataframe <- current_dataframe
  current_dataframe <- data.frame(
    HPO_ID <- tree$id,
    phenotype = tree$name,
    ancestors = paste(tree$is_a, collapse = ","),
    genes_associated = paste(c(tree$genes_assorted), collapse = ","),
    chrom_modifiers_associated = paste(c(tree$chrom_assorted), collapse = ","),
    p_value = ifelse(is.null(tree$fisher_test$p.value), NA, tree$fisher_test$p.value), 
    odds_ratio = ifelse(is.null(tree$fisher_test$estimate), NA, tree$fisher_test$estimate[[1]]), 
    confidence_interval = ifelse(is.null(tree$fisher_test$conf.int),NA,paste(c(tree$fisher_test$conf.int), collapse = "--"))
  )
  current_dataframe <- rbind(dataframe, current_dataframe)
  
  if(!all(is.na(tree$left))){current_dataframe <- tree_to_dataframe(tree$left, current_dataframe)
  }
  if(!all(is.na(tree$right))){current_dataframe <- tree_to_dataframe(tree$right, current_dataframe)
  }
  return(current_dataframe)
}

search_tree_dataframe <- tree_to_dataframe(hpo_bin_search_tree)
```